Relative Stable Maps in Genus One via Central Alignments

### 2019-01-29 GenusOneRelativeMaps.tex
More changes to section on DRC

So let us start with a compact type stable map which belongs to the double ramification locus:
\bcd
C \ar[r,"f"] \ar[d] & H \\
S=\Spec\kfield
\ecd
We will show that there exists a unique lift of this to a punctured stable map. Since $C$ has compact type, there is a unique choice of associated combinatorial type $\Delta$. This determines a minimal monoid $Q$ such that any punctured map enhancing the given stable map has base log structure given by:
\begin{align*}
Q \oplus \kfield^\st & \to \kfield \\
(q,\lambda) & \mapsto \begin{cases} \lambda \qquad \text{if } q =0 \\ 0 \qquad \text{otherwise} \end{cases}
\end{align*}
If we let $r$ denote the number of nodes of $C$ then there is a natural map $\N^r \to Q$ and by the universal property of minimality the punctured curve must be obtained by puncturing a log smooth curve pulled back along a log map
\begin{equation*} (\Spec\kfield,Q \oplus \kfield^\st) \to (\Spec\kfield,\N^r\oplus\kfield^\st)\end{equation*}
enhancing the given map $\N^r \to Q$ on the level of ghost sheaves. There is a $\Gm^r$ of moduli for such maps, but all these choices are canceled out by automorphisms of $(\Spec\kfield,Q\oplus\kfield^\st)$\footnote{(Navid) Here we are using the fact that $Q$ is in some sense bigger than $\N^r$.}. Therefore we obtain a unique log smooth curve:
\begin{equation*} (C,\Mcal_C) \to (\Spec\kfield,Q\oplus\kfield^\st).\end{equation*}
We must now construct the puncturing $\Mcal_{C^\circ}$. Let $\Pcal_C$ denote the divisorial log structure on $C$ with respect to the punctures. Then there is a unique log structure $\Mcal$ on $C$ such that
\begin{equation*} \Mcal_C = \Mcal \oplus_{\OO_C^\st} \Pcal_C \end{equation*}
where the pushout is taken in the category of monoid sheaves. We consider the sheaf:
\begin{equation*} \Ncal_C = \Mcal \oplus_{\OO_C^\st} \Pcal^{\gp}_C. \end{equation*}
By definition the puncturing $\Mcal_{C^\circ}$ should be an intermediate sheaf:
\begin{equation*} \Mcal_C \subseteq \Mcal_{C^\circ} \subseteq \Ncal_C. \end{equation*}
We will now construct this, whilst arguing that our construction gives the only puncturing possible. First we note that it is sufficient to define $\Mcal_{C^\circ}$ on a base for the topology on $C$. We choose a base such that for every base open set $U$:
\begin{enumerate}
\item $U$ contains at most one of the punctures of $C$;
\item if $U$ contains a puncture $p \in C$, then $U$ does not contain any of the nodes or markings of $C$, and $\OO_C(p)|_U$ is trivial.
\end{enumerate}
Note that these properties are stable under taking subsets, so to obtain such a base we simply choose an open cover with this property (which is certainly possible) and then consider the collection of all open subsets of our chosen cover.

Now let $U$ be such a base open set. If $U$ does not contain a puncture then we set
\begin{equation*} \Mcal_{C^\circ}(U) = \Mcal_C(U).\end{equation*}
If $U$ contains a puncture $p$ then by assumption $U \setminus \{ p\}$ consists of smooth, unmarked points, and hence we have
\begin{equation*} \Ncal_C|_U = \left( \pi^\st \Mcal_S|_U \right) \oplus_{\OO_U^\st} \left(\Pcal_p^{\gp}|_U \right) \end{equation*}
where $\Pcal_p$ is the divisorial log structure with respect to $p$. Now, we have $\Mcal_S = Q \oplus \kfield^\st$ from which it follows that:
\begin{equation*} \pi^*\Mcal_S|_U = \ul{Q} \oplus \OO_U^\st.\end{equation*}
Thus we see that
\begin{equation*} \Ncal_C|_U = \ul{Q} \oplus \Pcal_p^{\gp}|_U\end{equation*}
and similarly:
\begin{equation*} \Mcal_C|_U = \ul{Q} \oplus \Pcal_p|_U.\end{equation*}
We therefore have:
\begin{equation*} \Ncal_C(U) = Q \oplus \Pcal_p^{\gp}(U) \end{equation*}
Recall we are assuming that $\OO_C(p)|_U$ is trivial. Choosing a trivialisation, we obtain a function $s \in \OO_U(U)$ such that $p=\{s=0\}$. Then $\Pcal_p(U)$ consists of functions of the form $\lambda \cdot s^n$ for $n \in \N$ and $\lambda \in \OO_U^\st(U)$. Groupifying, we obtain:
\begin{equation*} \Pcal^{\gp}_p(U) = \left\{ \lambda \cdot s^k \, \vert \, \lambda \in \OO_U^\st(U), k \in \Z \right\} \end{equation*}
Recall that $p$ has attached to it a tangency order $\alpha_p \in \Z_{<0}$. If we let $\eta$ denote the generic point of the component of $C$ containing $p$, then we also have a pullback map $\ol\varphi_\eta \colon \N \to Q$.

We define $\Mcal_{C^\circ}(U) \subseteq \Ncal_C(U)$ to be the submonoid generated by $Q \oplus \Pcal_p(U)$ and:
\begin{equation*} (\ol\varphi_\eta(1),s^{\alpha_p}). \end{equation*}
This clearly satisfies the condition to be a puncturing. Moreover, by the prestability condition for puncturings, this is the only possibility for $\Mcal_{C^\circ}$. We thus have a punctured curve:
\begin{equation*} (C,\Mcal_{C^\circ}) \xrightarrow{\pi} (\Spec\kfield,Q\oplus\kfield^\st). \end{equation*}
(Note that the log enhancement of $\pi$ comes for free by taking the composition $\pi^\st \Mcal_S \to \Mcal_C \to \Mcal_{C^\circ}$.) It remains to construct the log enhancement of the map $f \colon C \to H$. The map $f^\flat \colon f^\st \Mcal_H \to \Mcal_{C^\circ}$ is already determined on the level of ghost sheaves. To enhance this to a map on the level of log structures entails choosing an isomorphism of line bundles:
\begin{equation*} f^* \OO_H(1) \cong \OO_C\left(\ol{f}^\flat(1)\right). \end{equation*}
The component-wise description of the right-hand side in \cite{RSPW} (together with the fact that $C$ is compact type, so its Picard group is just the product of the Picard groups of its components) shows that such an isomorphism exists if and only if the original stable map is in the double ramification locus.

@@

\subsection{Geometry of double ramification loci and their desingularisations} Thus, the spaces of punctured maps we encounter are always birational to double ramification loci. We will now discuss the geometry of these loci. In the genus zero case, the double ramification locus is equal to the entire space
\begin{equation*} \ol\Mcal^{\alpha}_{0,n}(H,d) = \M{0}{n}{H}{d} \end{equation*}
which is smooth. Thus integrals over $\Mpunct{0}{\alpha}{H}{d}$ are the same thing as integrals over $\M{0}{n}{H}{d}$.

The genus one case is more delicate: there is a ``main component'' of the double ramification locus, obtained as the closure of the locus where the source curve is smooth. This forms a divisor inside the main component of the moduli space of stable maps:
%\begin{equation*} \M

Fortunately, we
are only interested in the main component of the punctured space. As we did in \S \ref{}, we may define a logarithmic modification of
$\Mpunct{1}{\alpha}{H}{d}$ by imposing an alignment and factorisation condition.

\begin{lem}
Alignment and factorisation single out the closure of the locus of punctured maps with smooth source curve.
\end{lem}
\begin{proof}
 If it is true, we might be able to prove it via deformation theory. For this it might be useful to notice that there is a morphism $\Mpunct{g}{\alpha}{H}{d}\to \Mpunct{g}{\alpha}{\PP^N}{d}$ (because $H\subseteq\PP^N$ is strict), which is probably a closed immersion, and the loci with smooth source curve are isomorphic under this map. It seems plausible that the main omponent is log smooth over $B\Gm\subseteq[\Aaff^1/\Gm]$ (or the standard log point).
\end{proof}

\begin{lem}
The main component of punctured maps is birational to the main component of the DR locus.
\end{lem}
We probably want the latter statement to hold under imposing alignment and factorisation (keeping the log structure on the source curve, but forgetting that the map is log).

\begin{lem}
 Forgetting a marking $\fgt_i\colon\ol\Mcal_{1,n}^\alpha(H,d)^{\rm{main}} \to\VZ_{1,n-1}(H,d)$ is generically finite of degree $\alpha_i^2$ - except in one case.
\end{lem}
\begin{proof}
 Let us consider the case $d=0$ first. We may assume the source curve is smooth elliptic $E$. The map \[ \phi\colon E\to \Pic^0(E),\qquad x\mapsto\OO_E\left(\alpha_ix+\sum_{j=1,\ldots,\hat i\ldots,n}\alpha_jp_j\right)\]
 is an isogeny of degree $\alpha_i^2$. The locus of $(C,p_1,\ldots,\hat{p_i},\ldots,p_n)$ such that the kernel of $\phi$ contains one of the points $\{p_1,\ldots,\hat{p_i},\ldots,p_n\}$ is itself a double ramification locus inside $\oM_{1,n}$, hence non-generic - with one exception: namely, when $n=2$, $x=p_1$ is always a solution, but the curve lying above such point bubbles off a $\PP^1$. To see that it does not belong to the closure of the nice locus, notice that the rational function trivialising $\OO_C(\alpha_1p_1-\alpha_2p_2)$ should descend to the cusp, thus having a ramification point at the node; yet its ramification profile is determined by Riemann-Hurwitz, and it is entirely supported on $p_1$ and $p_2$.

@@

\begin{prop} $\VZ^{\operatorname{punct}}_{1,\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}(H,d_0)$ is irreducible of the expected dimension, and birational to the main component of the double ramification locus:
\begin{equation*} \ol{\Mcal}_{1,n_0+r}^{\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}(H,d_0) \subseteq \ol{\Mcal}_{1,n_0+r}(H,d_0) \end{equation*}\end{prop}

Now, the double ramification locus carries a natural virtual class, and one can write
\begin{equation*} \virt{\ol{\Mcal}_{1,n_0+r}^{\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}(H,d_0)} = \gamma \cap \virt{\ol{\Mcal}_{1,n_0+r}(H,d_0)} \end{equation*}
where $\gamma \in \Achow^1(\M{1}{n}{H}{d})$ is some explicit sum of tautological classes \cite{DRCBundle}. To identify the class of the main component, we split the virtual class of the double ramification locus as
\begin{equation*} \virt{\ol\Mcal^{\alpha}_{1,n}(H,d)} = [\ol\Mcal^{\alpha,\text{main}}_{1,n}(H,d)] + \sum_{\Gamma} \virt{\ol\Mcal^{\alpha,\Gamma}_{1,n}(H,d)} \end{equation*}
and this allows us to write
\begin{equation*} [\ol\Mcal^{\alpha,\text{main}}_{1,n}(H,d)] = \gamma^\prime \cap [\ol\Mcal^{\text{main}}_{1,n}(H,d)]\end{equation*}
where $\gamma^\prime$ is some other sum of tautological classes. This makes integrals over $\VZ^{\operatorname{punct}}_{1,\alpha^{(0)}\cup(-m_1,\ldots,-m_r)}(H,d_0)$ computable in terms of the reduced Gromov--Witten theory of $H\cong\PP^{N-1}$.

>>>>

So let us start with a compact type stable map which belongs to the double ramification locus:
\bcd
C \ar[r,"f"] \ar[d] & H \\
S=\Spec\kfield
\ecd
We will show that there exists a unique lift of this to a punctured stable map. Since $C$ has compact type, there is a unique choice of associated combinatorial type $\Delta$. This determines a minimal monoid $Q$ such that any punctured map enhancing the given stable map has base log structure given by:
\begin{align*}
Q \oplus \kfield^\st & \to \kfield \\
(q,\lambda) & \mapsto \begin{cases} \lambda \qquad \text{if } q =0 \\ 0 \qquad \text{otherwise} \end{cases}
\end{align*}
If we let $r$ denote the number of nodes of $C$ then by construction there is a natural map $\N^r \to Q$ and by the universal property of minimality our punctured curve must be obtained by puncturing a log smooth curve pulled back along a log map
\begin{equation*} (\Spec\kfield,Q \oplus \kfield^\st) \to (\Spec\kfield,\N^r\oplus\kfield^\st).\end{equation*}
This log map must enhance the given map $\N^r \to Q$ on the level of ghost sheaves. The moduli for such a log map is $\Gm^r$, but all these choices are canceled out by automorphisms of $(\Spec\kfield,Q\oplus\kfield^\st)$\footnote{(Navid) Here we are using the fact that $Q$ is in some sense bigger than $\N^r$.}. Thus we obtain, by pulling back the minimal log smooth curve over $(\Spec\kfield,\N^r \oplus\kfield^\st)$, a unique log smooth curve:
\begin{equation*} (C,\Mcal_C) \to (\Spec\kfield,Q\oplus\kfield^\st).\end{equation*}
We must now construct the puncturing $\Mcal_{C^\circ}$. Let $\Pcal_C$ denote the divisorial log structure on $C$ with respect to the punctures. It is easy to see that there is a unique log structure $\Mcal$ on $C$ such that
\begin{equation*} \Mcal_C = \Mcal \oplus_{\OO_C^\st} \Pcal_C \end{equation*}
where the pushout is taken in the category of monoid sheaves. We consider the sheaf:
\begin{equation*} \Ncal_C = \Mcal \oplus_{\OO_C^\st} \Pcal^{\gp}_C. \end{equation*}
By definition the puncturing $\Mcal_{C^\circ}$ should be an intermediate sheaf:
\begin{equation*} \Mcal_C \subseteq \Mcal_{C^\circ} \subseteq \Ncal_C. \end{equation*}
We will now construct this, and argue that our construction gives the only puncturing possible. First we note that it is sufficient to define $\Mcal_{C^\circ}$ on a base for the topology on $C$. We choose a base such that for every base open set $U$:
\begin{enumerate}
\item $U$ is connected;
\item $U$ contains at most one of the punctures of $C$;
\item if $U$ contains a puncture $p \in C$, then $U$ does not contain any of the nodes or markings of $C$, and $\OO_C(p)|_U$ is trivial.
\end{enumerate}
Note that these properties are stable under taking subsets, so to obtain such a base we simply choose an open cover with this property (which is certainly possible) and then take the collection of all open subsets of our chosen cover.

Now let $U$ be such an open set in our base. If $U$ does not contain a puncture then we set
\begin{equation*} \Mcal_{C^\circ}(U) = \Mcal_C(U).\end{equation*}
If $U$ contains a puncture $p$ then by assumption $U \setminus \{ p\}$ contains only smooth unmarked points, and hence we have
\begin{equation*} \Ncal_C|_U = \left( \pi^\st \Mcal_S|_U \right) \oplus_{\OO_U^\st} \left(\Pcal_p^{\gp}|_U \right) \end{equation*}
where $\Pcal_p$ is the divisorial log structure with respect to $p$. Since $\Mcal_S = Q \oplus \kfield^\st$ it follows that $\pi^*\Mcal_S|_U = \ul{Q} \oplus \OO_U^\st$. Thus we see that
\begin{equation*} \Ncal_C|_U = \ul{Q} \oplus \Pcal_p^{\gp}|_U\end{equation*}
and similarly:
\begin{equation*} \Mcal_C|_U = \ul{Q} \oplus \Pcal_p|_U.\end{equation*}
We therefore have:
\begin{equation*} \Ncal_C(U) = Q \oplus \Pcal_p^{\gp}(U) \end{equation*}
Recall we are assuming that $\OO_C(p)|_U$ is trivial. Choosing a trivialisation, we obtain a function $s \in \OO_U(U)$ such that $p=\{s=0\}$. Then $\Pcal_p(U)$ consists of functions of the form $\lambda \cdot s^n$ for $n \in \N$ and $\lambda \in \OO_U^\st(U)$. Groupifying, we obtain:
\begin{equation*} \Pcal^{\gp}_p(U) = \left\{ \lambda \cdot s^k \, \vert \, \lambda \in \OO_U^\st(U), k \in \Z \right\} \end{equation*}
(Indeed $\Pcal_p^{\gp}$ may be described as the sheaf of rational functions which are regular and invertible outside of $p$.) Recall that $p$ has attached to it a tangency order $\alpha_p \in \Z_{<0}$. If we let $\eta$ denote the generic point of the component of $C$ containing $p$, then we also have a pullback map $\ol\varphi_\eta \colon \N \to Q$. We define $\Mcal_{C^\circ}(U) \subseteq \Ncal_C(U)$ to be the submonoid generated by $Q \oplus \Pcal_p(U)$ and:
\begin{equation*} (\ol\varphi_\eta(1),s^{\alpha_p}). \end{equation*}
This clearly satisfies the condition to be a puncturing. Moreover, by the prestability condition for puncturings, this is the only possibility for $\Mcal_{C^\circ}$. We thus have a punctured curve:
\begin{equation*} (C,\Mcal_{C^\circ}) \xrightarrow{\pi} (\Spec\kfield,Q\oplus\kfield^\st). \end{equation*}
(Note that the log enhancement of $\pi$ is uniquely determined by taking the composition $\pi^\st \Mcal_S \to \Mcal_C \to \Mcal_{C^\circ}$.) It remains to construct the log enhancement of the map $f \colon C \to H$. The map $f^\flat \colon f^\st \Mcal_H \to \Mcal_{C^\circ}$ is already determined on the level of ghost sheaves. To enhance this to a map on the level of log structures entails choosing an isomorphism of line bundles:
\begin{equation*} f^* \OO_H(1) \cong \OO_C\left(\ol{f}^\flat(1)\right). \end{equation*}
The component-wise description of the right-hand side in \cite{RSPW} (together with the fact that $C$ is compact type, so its Picard group is just the product of the Picard groups of its components) shows that such an isomorphism exists if and only if the original stable map is in the double ramification locus.

@@

\subsection{Geometry of double ramification loci and their desingularisations} Thus, the spaces of punctured maps we encounter are always birational to double ramification loci. We will now discuss the geometry of these loci. In the genus zero case, the double ramification locus is equal to the entire space
\begin{equation*} \ol\Mcal^{\alpha}_{0,n}(H,d) = \M{0}{n}{H}{d} \end{equation*}
which is smooth. Thus tautological integrals over $\ol\Mcal^{\alpha}_{0,n}(H,d)$ are determined by the genus-zero Gromov--Witten theory of $H$.

The genus one case is more delicate: there is a ``main component'' of the double ramification locus, obtained as the closure of the locus where the source curve is smooth. This forms a divisor inside the main component of the moduli space of stable maps:
\begin{equation*} \ol\Mcal^{\alpha,\circ}_{1,n}(H,d) \subseteq \ol\Mcal^\circ_{1,n}(H,d). \end{equation*}
However, we know that $\M{1}{n}{H}{d}$ contains many other components besides $\ol\Mcal^\circ_{1,n}(H,d)$, some of which have excess dimension. Certainly the double ramification locus intersects these other components, and sometimes it does so in quite dramatic fashion. For instance, consider the locus where the source curve splits into a contracted elliptic component and a rational component containing all the markings. This locus forms an irreducible component of the moduli space, isomorphic to
\begin{equation*}\ol\Mcal_{1,1} \times \M{0}{n+1}{H}{d} \end{equation*}
which has excess dimension $N-2$. This entire component belongs to the double ramification locus, since the intersection-theoretic condition is entirely numerical here.

Thus we see that the double ramification locus can be quite badly-behaved in genus one. Fortunately, we will only be interested in the main component of the punctured space (which is birational to the main component of the double ramification locus). As we did in \S \ref{}, we may produce a logarithmic modification
\begin{equation*} \widetilde{\VZ}^{\operatorname{punct}}_{1,\alpha}(H,d) \to \Mpunct{1}{\alpha}{H}{d} \end{equation*}
by requiring our maps to be centrally aligned. We may then define a closed substack
\begin{equation*} \VZ^{\operatorname{punct}}_{1,\alpha}(H,d) \subseteq \widetilde{\VZ}^{\operatorname{punct}}_{1,\alpha}(H,d) \end{equation*}
by imposing a factorisation condition. These moduli spaces will appear in our recursive description of the boundary of $\VZ$ given later in this section.
\begin{lem} This process produces a birational modification
\begin{equation*}\VZ^{\operatorname{punct}}_{1,\alpha}(H,d) \to \ol\Mcal^{\operatorname{punct},\circ}_{1,\alpha}(H,d)\end{equation*}
of the main component of the moduli space of punctured maps (which, in turn, is birational to the main component of the double ramification locus).\end{lem}
\begin{proof}
 If it is true, we might be able to prove it via deformation theory. For this it might be useful to notice that there is a morphism $\Mpunct{g}{\alpha}{H}{d}\to \Mpunct{g}{\alpha}{\PP^N}{d}$ (because $H\subseteq\PP^N$ is strict), which is probably a closed immersion, and the loci with smooth source curve are isomorphic under this map. It seems plausible that the main omponent is log smooth over $B\Gm\subseteq[\Aaff^1/\Gm]$ (or the standard log point).
\end{proof}

%We probably want the latter statement to hold under imposing alignment and factorisation (keeping the log structure on the source curve, but forgetting that the map is log).
Integrals over the main component of the double ramification locus can be computed using the following lemma.
\begin{lem} The morphism which forgets a marking
\begin{equation*} \fgt_i\colon\ol\Mcal_{1,n}^{\alpha,\circ}(H,d)\to\ol\Mcal^{\circ}_{1,n-1}(H,d)\end{equation*}
is generically finite, of degree $\alpha_i^2$ (except in one special case, described in the proof).
\end{lem}
\begin{proof}
 Let us consider the case $d=0$ first. We may assume the source curve is smooth elliptic $E$. The map \[ \phi\colon E\to \Pic^0(E),\qquad x\mapsto\OO_E\left(\alpha_ix+\sum_{j=1,\ldots,\hat i\ldots,n}\alpha_jp_j\right)\]
 is an isogeny of degree $\alpha_i^2$. The locus of $(C,p_1,\ldots,\hat{p_i},\ldots,p_n)$ such that the kernel of $\phi$ contains one of the points $\{p_1,\ldots,\hat{p_i},\ldots,p_n\}$ is itself a double ramification locus inside $\oM_{1,n}$, hence non-generic - with one exception: namely, when $n=2$, $x=p_1$ is always a solution, but the curve lying above such point bubbles off a $\PP^1$. To see that it does not belong to the closure of the nice locus, notice that the rational function trivialising $\OO_C(\alpha_1p_1-\alpha_2p_2)$ should descend to the cusp, thus having a ramification point at the node; yet its ramification profile is determined by Riemann-Hurwitz, and it is entirely supported on $p_1$ and $p_2$.

@@
